The aims of science and an open science controversy at USEPA

### 2025-07-11 paper/paper.qmd
working on todos

abstract: "abstract"

@@

Something like Hicks' two views appears in philosophical debates over the argument from inductive risk. *[acceptance vs. belief]*

@@

@@

*[set blue L to like 40]*

@@

*[already using red/blue for health/science]*

@@

We argue that the results above can be fruitfully understood in terms of @HicksWhenVirtuesAre2022 account of two views of the aims of science.  Specifically, we claim that supporters deployed the narrow view, while opponents deployed the broad view.

@@

>>>>

abstract: |
    | *[update w/ conclusions]*
    | 
    | Environmental controversies are often framed in terms of "science vs. anti-science," with leftists and progressives typically claiming the side of science. However, their opponents often appeal to science as well. Hicks (2022) proposed that two rival views of the aims of science might be at play in these kinds of controversies.  On the "narrow view" science has purely epistemic aims; while the "broad view" takes science to also have pragmatic or practical aims. 
    | 
    | We argue that these rival views can be seen in the controversy over the proposed US EPA open data rule "Strengthening Transparency in Regulatory Science." The original justification for the rule appealed to the narrative of a science-wide replication crisis and the open science movement. In public comments, supporters of the rule likewise appealed to the narrow view of science, criticizing the titular "regulatory science" in epistemic rather than practical terms. While opponents of the rule appealed to the broad view, connecting science to the practical aim of protecting human health. 
    | 
    | We analyze approximately 21,000 public comments on the proposed rule, combining manual coding with machine learning to classify every comment as supporting or opposing, and using natural language processing (NLP) and text mining methods to compare how supporters and opponents talk about science in their comments.

@@

Something like Hicks' two views of the aims of science appears in philosophical debates over the argument from inductive risk. First, the argument itself treats the non-epistemic consequences of research as on a par with the epistemic consequences, similarly to the way the broad view does not prioritize epistemic aims over pragmatic ones. Next, in his objection to Rudner's original formulation of the argument, @LeviMustScientistMake1960 claimed that Rudner (and other proponents of the argument) relied on an assumption that "To choose to accept a hypothesis as true (or to believe that H is true) is equivalent to choosing to act on the basis of H relative to some specific objective P" (348), which he associated with decision theory and, later, behaviorism [@LeviSeriousnessMistakes1962]. Denying this assumption, @LeviMustScientistMake1960 distinguished the "theoretical" or epistemic objective of "accepting true answers to questions as true" from "practical" objectives (350). Similarly, @LeviSeriousnessMistakes1962 defends the view that the aim of scientific inquiry is to "replace doubt by true belief" (48).  This aim might include either "seek[ing] the truth and nothing but the truth" or, more broadly, "certain desirable properties in addition to truth" such as "simplicity, explanatory power," and presumably other epistemic virtues (48).  But, on either construal, "replac[ing] doubt by true belief" is "theoretical" or epistemic and "non-behavioral" (48). In other words, Levi objected to the argument from inductive risk by articulating the narrow view. *[Franco cites]*. Finally, in response to similar critiques of *[Douglas']* version of the argument *[eg, Lacey; Mayo?]*, critics of the value-free ideal adopted *[Cohen's?]* distinction between belief (in a claim as true) and acceptance (for practical purposes), explicitly scoping their arguments to acceptance *[cites]*. This move in the debate concedes Levi's rejection of the "behaviorist" identification of belief and acceptance — but also rejects the narrow view's purely epistemic aims for inquiry, placing defenders and critics of VFI on opposite sides of the two views of the aims of science.

@@

*[Gold Standard Science EO]*

@@

@@

@@

We argue that the results above can be fruitfully understood in terms of Hicks' -@HicksWhenVirtuesAre2022 account of two views of the aims of science.  Specifically, we claim that supporters deployed the narrow view, while opponents deployed the broad view.

@@

Yet another alternative account of this controversy would be in terms of the value-free ideal, understood as the view that non-epistemic values must not influence the "internal" or "core" stages of scientific inquiry — analyzing evidence, evaluating hypotheses — but may permissibly influence "earlier" and "later" stages of inquiry — such as posing research questions and utilizing research findings in policymaking. On this reading, supporters of ST perceive EPA/regulatory science as frequently violating the value-free ideal — perhaps environmentalist or "leftist" values are influencing the evaluation of hypotheses — while opponents do not.^[Thanks to [redacted]<!-- Soazig Le Bihan --> for suggesting a version of this objection. [Their] version did not tie the internal/external influence distinction to the value-free ideal.]

This account is not as much of an alternative to our aims-based account as it might seem. @HicksWhenVirtuesAre2022 argues that the narrow view is often assumed by defenders of the value-free ideal (VFI), and so for supporters of ST there's not much difference between "failing to achieve narrowly epistemic aims" and "violating VFI." However, for opponents of ST, this alternative account would require parsing their comments on ST into VFI's three-stage model of inquiry, and showing that they view non-epistemic values as influencing only the "external" stages (research questions and downstream applications). Concerns that ST would enable "political interference" — presidential appointees influencing the evaluation of hypotheses and conclusions reached by scientific reports — can be read as concerns about VFI violations. But other concerns would be much more difficult to parse in terms of the three-stage model assumed by VFI.

The VFI-based account of the ST controversy also does not fit well with empirical research on the relationship among VFI, cynicism about science, and trust in science among members of the general public in the US. @HicksDevelopingMeasuresPublic2024 find that members of the public generally agree with VFI, and the average score along a composite measure of "objectivity" is relatively high. But they also find that this measure does not predict trust in science. Instead a different composite measure, "cynicism," predicts trust in science,^[Specifically, a high score on the cynicism scale is a good predictor of *lower* trust in science.] and cynicism and objectivity are not correlated with each other. @HicksDevelopingMeasuresPublic2024 also connect the cynicism measure to the failure of science to achieve its aims. Below we discuss how supporters of ST often express significant cynicism towards EPA/regulatory science.
